A subjective Bayesian framework for synthesizing deep uncertainties in climate risk management

### 2022-07-05 papers/earths-future/submission.tex
incorporate rest of klaus handwritten suggestions

%% required packages
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}

% Figures
\usepackage{graphicx}
\usepackage[list=true]{subcaption}
\graphicspath{{../../plots/}{../../tikz/}{../../img/}}
\usepackage[section]{placeins} % require floats to appear in the section they are defined

@@

Many critical infrastructure services around the world are aging \cite{ho_dams:2017}.
Further, anticipated changes to regulation, economics and finance, patterns of population and infrastructure use, and climate challenge their ability to meet their design objectives \cite{doss-gollin_txtreme:2021,doss-gollin_fatalism:2020,chester_reliable:2020}.
The need to upgrade and expand infrastructure motivates the question: for which possible futures should infrastructure systems and components be designed?
The answer to this question depends in part on values: intrinsic trade-offs between safety, performance, cost, and other objectives depend on context and stakeholder preferences \cite{keller_management:2021} and are often regulated by statute or industry guidelines \cite{bruneau_multihazard:2017}.
For example, hospitals and critical infrastructure are generally designed to a higher standard of risk protection than ordinary buildings \cite{asce_7-10:2013}.
At the same time, performance depends upon future conditions, and so decisions about which scenario to design for are necessarily subject to implicit or explicit assumptions about the likelihood or possibility of different futures.

@@

The approaches currently used to estimate the \glspl{pdf} of climate hazards used in standard-setting and decision-making emphasize nominally objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies procedures for estimating flood frequency \cite{bulletin17c:2019} and \gls{noaa} Atlas 14 provides estimates of the intensity, duration, and frequency of extreme rainfall \cite{atlas14_texas:2018}.
Among several statistical assumptions these analyses make that have been recently called into question is stationarity \cite<the assumption that past and future hazard come from the same \gls{pdf}; see>[for a review]{merz_review:2014}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \cite{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}, and trends in many other hazards for other locations are consistent with observations \cite<see>[for a comprehensive summary]{ipcc_impacts:2022}.\footnote{Klaus comment TBD}
While some methods have been proposed for incorporating trends into these analyses \cite<see>[for a review]{Salas:2018ge}, these assume specific forms of nonstationarity which may not fully represent physical processes or true levels of uncertainty \cite{doss-gollin_robustadaptation:2019,Montanari:2014hl,Serinaldi:2015bq}.
Ultimately, the difficulty of incorporating nonstationarity into existing frameworks has led to continued reliance on stationarity despite recognition of the associated limitations as discussed in p.~2 of \citeA{bulletin17c:2019} and p.~A.4-42 of \citeA{atlas14_texas:2018}.

@@

Projecting nonstationary hazard is difficult because many future hazards depend on human decisions (\eg, future greenhouse gas emissions) or on physical processes that are poorly constrained by existing data \cite<\eg, collapse of the West Antarctic ice sheet;>[]{deconto_antarctica:2016}.
In other words, they are deeply uncertain \cite{keller_management:2021,walker_deep:2013,lempert_complex:2002}.
These deeply uncertain nonstationary hazards challenge not only the existing stationary estimates but, more fundamentally, the premise that objective estimates of future hazard exist and can be estimated empirically.

@@

We proceed as follows.
In \cref{sec:analysis} we present three formal decision analytic frameworks for analyzing an ensemble of \gls{slr} simulations, building through existing approaches for exploratory modeling (\cref{sec:analysis-explore}) and scenario analysis (\cref{sec:analysis-condition}) towards a formal Bayesian method for transparently synthesizing deep uncertainty through subjective prior beliefs (\cref{sec:analysis-synthesize}).
In \cref{sec:case-study} we describe the case study in detail.
Next in \cref{sec:results} we present results for each of the three decision lenses and discuss the advantages and limitations of each theoretical approach.
Finally in \cref{sec:conclusions} we discuss key findings, future research needs, and implications for policy and practice.

@@

Although exploratory modeling is a useful framework for understanding systems, there are many questions that it cannot answer.
For example, answering questions like ``what is the 95th percentile of metric $u$ under decision $x$'' or ``which decision minimizes expected damages'' or ``what is the probability of exceeding a critical threshold'' requires constructing estimators for the probability distribution over outcomes \cite<see>{schneider_scenarios:2002}.

@@

To illustrate the proposed decision analytic framework, we model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
Following \citeA{zarekarizi_suboptimal:2020}, we focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and approximate other model parameters as shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
  \item The decision vector $\vb{x}$ is comprised of discrete possible house heightenings ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$.
  \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=70$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
  \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{xlrm.pdf}
  \caption{
    Conceptual diagram of the considered example.
    A \acrfull{sow} consists of a description of the uncertain factors (red).
    We model a problem with a single lever (yellow), which is how high to elevate a house ($\Delta h$).
    For each \acrshort{sow} (red) and each value of $\Delta h$, the system model (blue) is used to calculate performance metrics (gray).
  }\label{fig:xlrm}
\end{figure}

@@

\begin{table}[h]
  \centering
  \caption{
    Subjective priors over \gls{slr} from 2022 to 2100, \ie $p_\mathrm{belief}(\psi)$.
    The name of the distribution, the parameters of the Gamma distribution with shape $\alpha$ and scale $\theta$, and the 2.5, 25, 50, 75, and 97.5th percentiles (values in \si{ft}).
  }\label{tab:slr-priors}
  \begin{tabular}{llllllll}
    \toprule
    Name          & \multicolumn{2}{c}{Parameters} & \multicolumn{5}{c}{Percentiles (in \si{ft})}                                     \\
    \cmidrule(lr){2-3}
    \cmidrule(lr){4-8}
                  & $\alpha$                       & $\theta$                                     & 2.5  & 25.0 & 50.0 & 75.0 & 97.5  \\
    \midrule
    Slow SLR      & 1.75                           & 0.50                                         & 0.08 & 0.39 & 0.72 & 1.19 & 2.57  \\
    Uncertain SLR & 1.75                           & 1.25                                         & 0.21 & 0.98 & 1.79 & 2.97 & 6.41  \\
    Rapid SLR     & 3.50                           & 1.25                                         & 1.06 & 2.66 & 3.97 & 5.65 & 10.01 \\
    \bottomrule
  \end{tabular}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{tradeoffs-by-rcp}
  \caption{
    Each probabilistic model or scenario leads to a different estimate of the Pareto frontier.
    For emphasis, we highlight three representative models: the Brick Slow model \protect\cite{wong_brick0.2:2017} under \gls{rcp} 2.6, the K14 \protect\cite{kopp_probabilistic:2014} model under \gls{rcp} 6.0 and the DP16 model \protect\cite{deconto_antarctica:2016,kopp_evolving:2017} under \gls{rcp} 8.5.
    (A): trade-off between up-front cost (which is a monotonic function of height increase) and expected lifetime costs.
    (B): trade-off between up-front cost and lifetime expected damages (eq.~\ref{eq:led}).
    Light gray lines show estimates for all 16 models (four \gls{rcp} scenarios times four physical parameterizations) considered.
    Colored lines highlight three representative models for emphasis.
  }\label{fig:tradeoffs-by-rcp}
\end{figure}

@@

This approach is, in a sense, another form of exploratory modeling: instead of considering a very large ensemble of \glspl{sow}, we consider a much smaller set of probabilistic models.
This approach is attractive because it allows modelers to focus on their domain expertise (\eg, the response of ice sheets and global sea levels to a particular climate future).
However, conditioning simulations on a set of climate futures and physical models presents what we term ``the multiple \gls{pdf} problem'' because it leaves decision makers with many \glspl{pdf} to choose from and hence many trade-off curves to navigate.
The multiple \gls{pdf} problem has also been shown in other contexts.
For example, \citeA{sharma_rcp:2021} model the reliability of stormwater infrastructure under different climate models and downscaling methods, finding diverging estimates of future rainfall hazard, even under a single \gls{rcp} scenario.
Similarly, \citeA{wong_nola:2017} construct 18 probability distribution functions for future flood risk in New Orleans, considering multiple models for ice sheet dynamics and storm surge and multiple \gls{rcp} scenarios.
Although this scenario-conditional analysis is appropriate for understanding differences between models, its key limitation is that it \textbf{places the burden for deciding which model to design for onto the end user.}

@@

\subsection{Implications for research and practice}

@@

\subsection{Conclusions and implications}

The proposed scenario weighting framework can be applied to inform critical challenges in climate risk management.
An obvious area of application is to the design of infrastructure.
For example, much of the stormwater infrastructure in the United States is inadequate for current climates \cite{lopez-cantu:2018}, let alone future changes.
Yet upgrading this infrastructure is costly and subject to large uncertainties between rainfall models \cite{sharma_rcp:2021} and \gls{rcp} scenarios.
Similarly, decisions like  levee heightening \cite{garner_slrise:2018,oddo_coastal:2017,vandantzig_dike:1956} and sea wall design (as discussed in the Introduction) are subject to deep uncertainties in sea level rise.
Investments in water resources planning and management depend on assumptions of future water demand, availability, and technologies \cite{trindade_deeplyuncertainpathways:2019}.
And analyses of climate change mitigation options, such as estimates of the social cost of pollutants \cite{errickson_methane:2021} or cost-minimizing energy transition pathways, are conditional on probabilistic models for inputs like technology prices and population.

>>>>

%% required packages
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.

% Figures
\usepackage{graphicx}
\graphicspath{{../../plots/}{../../tikz/}{../../img/}}
\usepackage[section]{placeins} % require floats to appear in the section they are defined

@@

Many critical infrastructure services around the world are aging \cite<\eg,>{ho_dams:2017}.
Moreover, anticipated changes to regulation, economics and finance, patterns of population and infrastructure use, and climate challenge their ability to meet their design objectives \cite{doss-gollin_txtreme:2021,doss-gollin_fatalism:2020,chester_reliable:2020}.
The need to upgrade and expand infrastructure motivates the question: for which possible futures should infrastructure systems and components be designed?
The answer to this question depends in part on values: intrinsic trade-offs between safety, performance, cost, and other objectives depend on context and stakeholder preferences \cite{keller_management:2021} and are often regulated by statute or industry guidelines \cite{bruneau_multihazard:2017}.
For example, hospitals and critical infrastructure are generally designed to a higher standard of risk protection than ordinary buildings \cite{asce_7-10:2013}.
At the same time, performance depends upon future conditions, and so decisions about which scenario to design for are necessarily subject to implicit or explicit assumptions about the likelihood or possibility of different futures.

@@

The approaches currently used to estimate the \glspl{pdf} of climate hazards used in standard-setting and decision-making emphasize nominally objective methods that can be applied consistently across locations.
For example, \gls{usgs} Bulletin 17C specifies procedures for estimating flood frequency \cite{bulletin17c:2019} and \gls{noaa} Atlas 14 provides estimates of the intensity, duration, and frequency of extreme rainfall \cite{atlas14_texas:2018}.
Among several statistical assumptions these analyses make that have been recently called into question is stationarity \cite<the assumption that past and future hazard come from the same \gls{pdf}; see>[for a review]{merz_review:2014}.
For example, clear trends in extreme rainfall are apparent across much of southeastern Texas \cite{fagnant_spatiotemporal:2020,nielsen-gammon_txrainfall:2020}, and trends in many other hazards for other locations are consistent with observations \cite<see>[for a comprehensive summary]{ipcc_impacts:2022}.
While some methods have been proposed for incorporating trends into these analyses \cite<see>[for a review]{Salas:2018ge}, these assume specific forms of nonstationarity which may not fully represent physical processes or true levels of uncertainty \cite{doss-gollin_robustadaptation:2019,Montanari:2014hl,Serinaldi:2015bq}.
Ultimately, the difficulty of incorporating nonstationarity into existing frameworks has led to continued reliance on stationarity despite recognition of the associated limitations as discussed in p.~2 of \citeA{bulletin17c:2019} and p.~A.4-42 of \citeA{atlas14_texas:2018}.

@@

Projecting nonstationary hazard is difficult because many future hazards depend on human decisions (\eg, future greenhouse gas emissions) or on physical processes that are poorly constrained by existing data \cite<\eg, collapse of the West Antarctic ice sheet;>[]{deconto_antarctica:2016}.
In other words, they are deeply uncertain \cite{keller_management:2021,walker_deep:2013,lempert_complex:2002,haasnoot_sealevelrise:2021}.
These deeply uncertain nonstationary hazards challenge not only the existing stationary estimates but, more fundamentally, the premise that objective estimates of future hazard exist and can be estimated empirically.

@@

We proceed as follows.
In \cref{sec:analysis} we present three formal decision analytic frameworks for analyzing an ensemble of \gls{slr} simulations, building through existing approaches for exploratory modeling (\cref{sec:analysis-explore}) and scenario analysis (\cref{sec:analysis-condition}) towards a formal Bayesian method for transparently synthesizing deep uncertainty through subjective prior beliefs (\cref{sec:analysis-synthesize}).
In \cref{sec:case-study} we describe the case study in detail.
Next in \cref{sec:results} we present results for each of the three decision lenses and discuss the advantages and limitations of each theoretical approach.
In \cref{sec:caveats} we discuss limitations of the study and future research needs.
Finally in \cref{sec:conclusions} we discuss key findings and implications for policy and practice.

@@

Although exploratory modeling is a useful framework for understanding systems, there are many questions that it cannot answer.
For example, answering questions like ``what is the 95th percentile of metric $u$ under decision $x$'' or ``which decision minimizes expected damages'' or ``what is the probability of exceeding a critical threshold'' requires constructing estimators for the probability distribution over outcomes \cite<see>[for a general discussion]{schneider_scenarios:2002}.

@@

To illustrate the proposed decision analytic framework, we model a one-time decision of whether to elevate a house, and if so by how much (\cref{fig:xlrm}).
Following the approach outlined in \citeA{zarekarizi_suboptimal:2020}, we focus on a case study of a \emph{hypothetical} house in Norfolk, VA.
For interpretability, we focus on deep uncertainty in \gls{msl} and approximate other model parameters as shallow uncertainties as shown in \cref{tab:uncertainties}.
We use the notation developed in the previous section to describe the case study.
Specifically,
\begin{enumerate}
  \item The decision vector $\vb{x}$ is comprised of discrete possible house heightenings ($\Delta h$); we consider $\Delta h = \qty{\SI{0}{ft}, \SI{0.25}{ft}, \ldots, \SI{12}{ft}}$.
  \item The \glspl{sow} describe annual time series of \gls{msl} over the $T=70$ year house lifetime so $\vb{s} \in \mathbb{R}^T$
  \item The system model $f$ quantifies up-front costs and lifetime expected damages, given a decision $x_i$ and \gls{sow} $s_j$, by integrating economic and engineering damage models over a probability distribution for storm surge.
\end{enumerate}
In the remainder of this section we describe data sources and treatment of \gls{slr} (\cref{sec:case-slr}), storm surge (\cref{sec:case-surge}), damages and metrics (\cref{sec:case-metrics}), and finally the subjective priors $p_\mathrm{belief}$ used to apply the re-weighting method described in \cref{sec:analysis-synthesize} to this case study (\cref{sec:case-priors}).

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{xlrm.pdf}
  \caption{
    Conceptual diagram of the considered example.
    A \acrfull{sow} consists of a description of the uncertain factors (red).
    We model a problem with a single lever (yellow), which is how high to elevate a house ($\Delta h$).
    For each \acrshort{sow} (red) and each value of $\Delta h$, the system model (blue) is used to calculate performance metrics (gray).
    We also compute a third metric, expected lifetime costs, which is the sum of up-front costs and lifetime expected damages.
  }\label{fig:xlrm}
\end{figure}

@@

\begin{table}[ht]
  \centering
  \caption{
    Subjective priors over \gls{slr} from 2022 to 2100, \ie $p_\mathrm{belief}(\psi)$.
    The name of the distribution, the parameters of the Gamma distribution with shape $\alpha$ and scale $\theta$, and the 2.5, 25, 50, 75, and 97.5th percentiles (values in \si{ft}).
  }\label{tab:slr-priors}
  \begin{tabular}{llllllll}
    \toprule
    Name          & \multicolumn{2}{c}{Parameters} & \multicolumn{5}{c}{Percentiles (in \si{ft})}                                     \\
    \cmidrule(lr){2-3}
    \cmidrule(lr){4-8}
                  & $\alpha$                       & $\theta$                                     & 2.5  & 25.0 & 50.0 & 75.0 & 97.5  \\
    \midrule
    Slow SLR      & 1.75                           & 0.50                                         & 0.08 & 0.39 & 0.72 & 1.19 & 2.57  \\
    Uncertain SLR & 1.75                           & 1.25                                         & 0.21 & 0.98 & 1.79 & 2.97 & 6.41  \\
    Rapid SLR     & 3.50                           & 1.25                                         & 1.06 & 2.66 & 3.97 & 5.65 & 10.01 \\
    \bottomrule
  \end{tabular}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{tradeoffs-by-rcp}
  \caption{
    Each probabilistic model or scenario leads to a different estimate of the Pareto frontier.
    For emphasis, we highlight three representative models: the Brick Slow model \protect\cite{wong_brick0.2:2017} under \gls{rcp} 2.6, the K14 \protect\cite{kopp_probabilistic:2014} model under \gls{rcp} 6.0 and the DP16 model \protect\cite{deconto_antarctica:2016,kopp_evolving:2017} under \gls{rcp} 8.5.
    (A): trade-off between up-front cost (which is a monotonic function of height increase) and expected lifetime costs.
    (B): trade-off between up-front cost and lifetime expected damages (eq.~\ref{eq:led}).
    Light gray lines show estimates for all 16 models (four \gls{rcp} scenarios times four physical parameterizations) considered.
    Colored lines highlight three representative models for emphasis.
    Gray lines and the smiling face in the lower left corner indicate the direction of preference.
  }\label{fig:tradeoffs-by-rcp}
\end{figure}

@@

This approach is, in a sense, another form of exploratory modeling: instead of considering a very large ensemble of \glspl{sow}, we consider a much smaller set of probabilistic models.
This approach is attractive because it allows modelers to focus on their domain expertise (\eg, the response of ice sheets and global sea levels to a particular climate future).
However, conditioning simulations on a set of climate futures and physical models presents what we term ``the multiple \gls{pdf} problem'' because it leaves decision makers with many \glspl{pdf} to choose from and hence many trade-off curves to navigate.
The multiple \gls{pdf} problem has also been shown in other contexts.
For example, \citeA{sharma_rcp:2021} model the reliability of stormwater infrastructure under different climate models and downscaling methods, finding diverging estimates of future rainfall hazard, even under a single \gls{rcp} scenario.
Similarly, \citeA{wong_nola:2017} construct 18 probability distribution functions for future flood risk in New Orleans, considering multiple models for ice sheet dynamics and storm surge and multiple \gls{rcp} scenarios and \citeA{haasnoot_sealevelrise:2021} identify global adaptation needs for different scenarios of \gls{slr}.
Although this scenario-conditional analysis is appropriate for understanding differences between models, its key limitation is that it \textbf{places the burden for deciding which model to design for onto the end user.}

@@

\section{Discussion and research needs}\label{sec:caveats}

@@

The proposed scenario weighting framework can be applied to inform critical challenges in climate risk management.
An obvious area of application is to the design of infrastructure.
For example, much of the stormwater infrastructure in the United States is inadequate for current and anticipated future climates \cite{lopez-cantu:2018}.
Yet upgrading this infrastructure is costly and subject to large uncertainties between rainfall models \cite{sharma_rcp:2021} and \gls{rcp} scenarios.
Similarly, decisions like  levee heightening \cite{garner_slrise:2018,oddo_coastal:2017,vandantzig_dike:1956} and sea wall design (as discussed in the Introduction) are subject to deep uncertainties in sea level rise.
Investments in water resources planning and management depend on assumptions of future water demand, availability, and technologies \cite{trindade_deeplyuncertainpathways:2019}.
And analyses of climate change mitigation options, such as estimates of the social cost of pollutants \cite{errickson_methane:2021} or cost-minimizing energy transition pathways, are conditional on probabilistic models for inputs like technology prices and population.
